\documentclass[12pt]{article}
\usepackage{enumitem}

\begin{document}
\section{Task 1}
Identify ambiguities or omissions in the requirements for a part of a ticket issuing system.  \\ Consider both functional and non-functional requirements.  \\ 
“A ticket issuing system is intended to automate the sale of rail tickets. Users select their destination, and input a credit card and a personal identification number. The aril ticket is issued and the credit card account charged with its cost. When the user presses the start button, a menu display of potential destinations is activated along with a message to the user to select a destination. Once a destination has been selected, users are requested to input their credit card. Its validity is checked and the user is then requested to input a personal identifier. When the credit transaction has been validated, the ticket is issued.”  


\section{Task 2}

A system that supports taxi agency has to be prepared: 

\begin{enumerate}[label=\Alph*)]
\item Identify stakeholders 

\item  Select one stakeholder and prepare an interview (write at least 6 important questions) 

\item Write specification of at least 3 different non-functional requirements of the system 

\item Write specification of at least 3 different non-functional requirements of the system 
 Estimate priorities using one of methods: cumulative votes with buying ideas, or 3-stage voting with weights (e.g. 9-5-3). For example, five functional requirements have been identified, a group of 4 stakeholders is available. Take assumptions about stakeholders, if necessary. 
\end{enumerate}


\section{Task 3}

In a project, problems with a staff recruitment are quit probable (50\%) and may cause serious effects (10 weeks delay). A hardware to be tested could be not available on time (25\%, 4 weeks delay). Calculate the global risk.  \\ 

We can hire another person for 20 weeks (cost \$200/week), which lower the staff risk to 30\%.  \\   
Or we can borrow a hardware in another company for 500\$ avoiding the second risk. 
Compare the final costs in both situations. Justify any business decision (to do nothing, or to cope with one of the risks, or both). 

A fine of a late project is about 1000\$ per week.  


\section{Task 4}
In the Cocomo model the number of persons involved in the project is the same during the whole project realization time. True, False or elsewhere? Justify the answer. 

\end{document}